Practical consistency management for geographically distributed MMOG servers

### 2011-05-03 trunk/textos/optpaxos/draft.tex
liveness achieved, I think

A game is composed of a set of objects. The game state is defined by the individual states of each one of its objects. We assume that the game objects are partitioned among different servers. Since objects have a location in the game, one way to perform this partitioning is by zoning the virtual world of the game. Each partition consists of a set of objects of the game and the server responsible for them is their \emph{coordinator}. As partitions represent physical regions in the game world, we define \emph{neighbor} regions as those which share a border with each other. We consider that each server is replicated, thus forming several groups which consist of the coordinator and its replicas. Therefore, for each region of the game world, there is a group assigned to it. This way, a group is said to coordinate an object when the group's coordinator is the coordinator for that object. Finally, groups are called neighbors when they are assigned to neighbor regions.

Each player may send his command to one of the replicas, instead of the coordinator, if that provides a lower delay between the issuing of a command and its delivery. In the case of avatar based virtual environments, and as an avatar is usually also an object, the server to which a player is connected is his avatar's coordinator. A command $C = \{ c_1, c_2, ... \}$ is composed of one or more subcommands, one subcommand per object it affects. We refer to the set of objects affected by command $C$ as $obj(C)$. Also, we refer to the objects coordinated by a server $S$ as $obj(S)$. Finally, we define $obj(C,S)$ = \{$o$ : $o \in obj(C)$ and $o \in obj(S)$\}.

@@

To ensure reliability despite server crashes, each server is replicated using state-machine replication, implemented with Paxos. Each player may send its commands to any of his server's replicas -- which then forwards it to the server. Each command is assigned a timestamp and executed against objects in timestamp order. We implement this by using a logical clock in each server group, considering that each group consists of a server and its replicas. Guaranteeing that the same set of commands is executed by the respective affected objects in the same order provides the level of consistency we are seeking. Therefore, the challenge is how to assign timestamps to commands such that consistency is not violated and commands are not discarded due to stale timestamp values. There are three cases to consider:

However, providing such level of consistency may prove to be costly in an MMOG context, since there may be several communication steps between the sending of a command and its in-order atomic delivery. For this reason, we also use an \emph{optimistic} delivery for the commands of the players. This delivery, although not guaranteeing that all replicas will receive all commands, or that they will be delivered in the same order, is designed to have a fairly low latency, counting from when each command is sent until it is delivered.

The final -- conservative, fault tolerant, but costly in terms of communication steps for each message -- delivery order should be as close as possible to the optimistic one, so that no roll-backs would be necessary. To explain how this works, we first should understand the idea behind the optimistic delivery.

@@

After applying the provisory timestamp to $C$, the server $S$ immediately forwards it to all the other servers involved, including servers in other groups. A server is involved with a command $C$ when one of the objects in $obj(C)$ is coordinated by that server's group. Then, $S$ puts $C$ on a list, where it stays until $now>C.pTS+w(S)$, which means that $C$ has been in that list for a time longer than the defined wait window. In the meantime, other commands, sent from other clients and forwarded by other servers, may have been received and also inserted in that list, always sorting by the provisory timestamp of these commands. If $w(S)$ has been correctly estimated as the highest communication delay between the server and each of the other servers in its group or in its neighboring groups, and no message was lost, than all the commands that were supposed to be delivered before $C$ have necessarily been received already. If the same has occurred for all the servers, then all of them have received all the commands, and can deliver them in the same order of $pTS$.

@@

Instead of having the conservative delivery done completely independently from the optimistic one, we can actually use the latter as a hint for the final delivery order. Since the optimistic delivery should be fast compared to the conservative one, waiting for it should not decrease the system performance significantly. Also, if we wait a short period longer, we can avoid a rollback later caused by mismatches between the two delivery orders, which is not desirable. The basic idea is, then, to pick the optimistic delivery order seen at the coordinator of each group.

The complicating factor is the possibility of a command affecting the state of objects in more than one group. So, all involved groups must somehow agree regarding the delivery order of these commands. However, as in most MMOGs objects interact only with other objects which are nearby, we can exploit the fact that the groups addressed by a single command are usually neighbors. We do this by defining \emph{barriers} for multicast, such that $barrier(G_{send},G_{recv})$ = $t$ means that the group $G_{send}$ promised that no commands with timestamp lower than $t$ would be sent to group $G_{recv}$ anymore. When a server $S$ has received all the barrier values from its neighbours, and they are all greater than a value $t$, then $S$ knows that no more commands are coming from other groups and that, once the local ordering is done, all the commands with timestamp up to $t$ can be conservatively delivered. Besides, a barrier is sent along with the bundle of all messages with timestamp greater than the last previous barrier sent from $G_{send}$ to $G_{recv}$, so that when a server has received a barrier, it means that it knows all the messages sent until the time value stored in that barrier.

@@

  In this case, when $C$ is optimistically delivered by the coordinator, it starts a Paxos consensus instance within its group, having not only its own group's servers as learners, but also those of the neighbor groups affected by $C$. The proposed value is the command $C$ itself, including $C.pTS$. When the value is decided and learnt, the servers of each of the other groups $G'$ adjusts $barrier(G, G')$ to $C.pTS$ and the coordinators of all the involved groups insert it into the respective \textit{consPending} lists.

@@

A more formal description of the protocol is given in Algorithm \ref{algorithm:optpaxos}. Note that, by abuse of notation, we have `server $s \in C.dst$' instead of `server $s$~: $\exists$ group $G~\in~C.dst$ $\wedge$ $s~\in~G$. Also, we consider that three primitives are given: $getTime()$, which returns the current value of the local wallclock; $Propose(val, learners)$, which initiates a Paxos \cite{paxosref} instance proposing the value $val$ with the local group as the group of acceptors and $learners$ as the group of learners; and also $Learn(val)$, which is called when a paxos instance achieves consensus. The $Learn(val)$ is called for all the processes which are supposed to learn a given proposed value once it is decided by a Paxos instance. For the sake of simplicity, we assume that, in a group, the values are learnt by consensus in the same order in which the consensus instances were initiated. This can be easily done with a sequence number given by the group's coordinator when initiating each consensus instance.

Moreover, each server contains an \textit{optPending} list which contains the commands waiting to be \textit{OPT-Delivered} (optimistically delivered). The \mbox{\textit{consPending}} list contains the commands ready to be \textit{CONS-Delivered} (conservatively delivered), but which may be waiting for the clearance from the neighbor groups. Also, each command $cmd$ has at least three fields: $dst$, which is the list of groups which contain objects affected by it; $src$, which is the group to which the client who issued it is connected; and $pTS$, the value of the wallclock of a server when it received the command from the client.

\begin{algorithm}
\begin{distribalgo}[1]
\STATE \underline{This part of the algorithm is executed by every server $s$:}
\STATE
\INDENT{\textit{When $s$ receives a command $cmd$ from a client}}
  \STATE $cmd.pTS \leftarrow getTime()$  
  \COMMENT{current wallclock value as the timestamp of $cmd$}
  \INDENT{\textbf{for all} $s' \in cmd.dst$ \textbf{do}}
    \STATE $send(s', cmd)$    
    \COMMENT{send optimistically $cmd$ to all involved servers}
  \ENDINDENT
  \STATE $optPending \leftarrow optPending$ $\cup$ $\{cmd\}$
\ENDINDENT
\STATE
\INDENT{\textit{When a server $s$ receives a command $cmd'$ from another server}}
  \IF {$cmd'.pTS < getTime() - w(s)$}
    \STATE discard $cmd'$
    \COMMENT{late commands probably lead to out-of-order delivery}
  \ELSE
    \STATE $optPending \leftarrow optPending$ $\cup$ $\{cmd'\}$
  \ENDIF
\ENDINDENT
\STATE
\INDENT{\textit{When $\exists c \in optPending$ : $getTime() > c.pTS + w(s)$ $\wedge$\\$\nexists c' \in optPending: c'.pTS < c.pTS$}}
  \STATE OPT-Deliver(c)  
  \INDENT{\textbf{for all} $cli \in clients(s)$ \textbf{do}}
    \STATE $send(cli, \{c, \text{`optimistic'}\})$    
    \COMMENT{s sends $c$ to its clients, warning it is optimistic}
  \ENDINDENT
  \STATE $optPending \leftarrow optPending \setminus \{c\}$
  \IF {$s$ is a coordinator $\wedge$ $|c.dst| = 1$}
    \STATE $consPending \leftarrow consPending$ $\cup$ $\{c\}$
  \ENDIF  
\ENDINDENT
\STATE
\INDENT{\textit{When} $Learn(\{c, \text{`deliver'}\})$}
  \STATE CONS-Deliver(c)
  \INDENT{\textbf{for all} $cli \in clients(s)$ \textbf{do}}
    \STATE $send(cli, {c, \text{`final'}})$    
    \COMMENT{s sends $c$ to its clients, telling it is the final order}
  \ENDINDENT
  \IF {$s$ is a coordinator}
    \STATE $consPending \leftarrow consPending \setminus \{c\}$
  \ENDIF  
\ENDINDENT
\STATE
\STATE
\STATE \underline{The next part is executed only by the coordinator $s_c$ of each group $g$:}
\STATE
\INDENT{\textit{When $\exists c \in optPending$ : $getTime() > c.pTS + w(s_c)$ $\wedge$ \\ $|c.dst| > 1 \wedge c.src = g \wedge \nexists c' \in optPending : c'.pTS < c.pTS$}}
  \STATE $Propose(\{c, \text{`multicast'}\}, c.dst)$
  \COMMENT{saving c in the acceptors before multicasting}
\ENDINDENT
\STATE
\INDENT{\textit{When} $Learn(\{c, \text{`multicast'}\})$}
  \STATE $consPending \leftarrow consPending$ $\cup$ $\{c\}$ 
  \IF {$c.src \neq g$}
    \STATE $barrier(c.src, g) \leftarrow c.pTS$
  \ENDIF  
\ENDINDENT
\STATE
\INDENT{\textit{When $\exists c \in consPending$ : \\ $c.pTS < getTime() - w(s_c)$ $\wedge$ \\ $\forall g' : g'$ is a neighbor of $g$, $c.pTS < barrier(g',g)$ $\wedge$ \\ $\nexists c' \in consPending : c'.pTS < c.pTS$ }}
  \STATE $Propose(\{c, \text{`deliver'}\}, g)$
\ENDINDENT

@@

>>>>

A game is composed of a set of objects. The game state is defined by the individual states of each one of its objects. We assume that the game objects are partitioned among different servers. Since objects have a location in the game, one way to perform this partitioning is by zoning the virtual world of the game. Each partition consists of a set of objects of the game and the server responsible for them is their \emph{coordinator}. As partitions represent physical regions in the game world, we define \emph{neighbor} regions as those which share a border with each other. We consider that each server is replicated, thus forming several groups which consist of the coordinator and its replicas. Therefore, for each region of the game world, there is a group assigned to it. This way, a group is said to coordinate an object when the group's coordinator is the coordinator for that object. Finally, groups are called neighbors when they are assigned to neighbor regions. From now on, the word `server' will be used for any kind of server, be it a coordinator or a replica.

Each player may send his command to one of the servers -- which might not be the group's coordinator, if that provides a lower delay between the issuing of a command and its delivery. In the case of avatar based virtual environments, and as an avatar is usually also an object, the server to which a player is connected belongs to the group of his avatar's coordinator. A command $C = \{ c_1, c_2, ... \}$ is composed of one or more subcommands, one subcommand per object it affects. We refer to the set of objects affected by command $C$ as $obj(C)$. Also, we refer to the objects coordinated by a server $S$ as $obj(S)$. Finally, we define $obj(C,S)$ = \{$o$ : $o \in obj(C)$ and $o \in obj(S)$\}.

@@

To ensure reliability despite server crashes, each server is replicated using state-machine replication, implemented with Paxos. Each player sends his commands to the server to which he is connected, which then forwards it to the coordinator. Each command is assigned a timestamp and executed against objects in timestamp order. We implement this by using a logical clock in each server group. Guaranteeing that the same set of commands is executed by the respective affected objects in the same order provides the level of consistency we are seeking. Therefore, the challenge is how to assign timestamps to commands such that consistency is not violated and commands are not discarded due to stale timestamp values.

However, providing such level of consistency may prove to be costly in an MMOG context, since there may be several communication steps between the sending of a command and its atomic delivery. For this reason, we also use an \emph{optimistic} delivery for the commands of the players. This delivery, although not guaranteeing that all replicas will receive all commands, or that they will be delivered in the same order, is designed to have a fairly low latency, counting from when each command is sent until it is delivered.

The final -- conservative, fault tolerant, but costly in terms of communication steps for each message -- delivery order should be as close as possible to the optimistic one, so that no rollbacks would be deemed necessary. To explain how this works, we first should understand the idea behind the optimistic delivery.

@@

After applying the provisory timestamp to $C$, the server $S$ immediately forwards it to all the other servers involved, including servers in other groups. A server is involved with a command $C$ when one of the objects in $obj(C)$ is coordinated by that server's group. Then, $S$ puts $C$ on a list (\mbox{\textit{optPending}}), where it stays until $now>C.pTS+w(S)$, which means that $C$ has been in that list for a time longer than the defined wait window. In the meantime, other commands, sent from other clients and forwarded by other servers, may have been received and also inserted in that list, always sorting by the provisory timestamp of these commands. If $w(S)$ has been correctly estimated as the highest communication delay between the server and each of the other servers in its group or in its neighboring groups, and no message was lost, than all the commands that were supposed to be delivered before $C$ have necessarily been received already. If the same has occurred for all the servers, then all of them have received all the commands, and can deliver them in the same order of $pTS$.

@@

Instead of having the conservative delivery done completely independently from the optimistic one, we can actually use the latter as a hint for the final delivery order. Since the optimistic delivery should be fast when compared to the conservative one, waiting for it should not decrease the system performance significantly. Also, if we wait a short period longer, we can avoid a rollback later caused by mismatches between the two delivery orders, which is not desirable. The basic idea is, then, to pick the optimistic delivery order seen at the coordinator of each group.

The complicating factor is the possibility of a command affecting the state of objects in more than one group. So, all involved groups must somehow agree regarding the delivery order of these commands. However, as in most MMOGs objects interact only with other objects which are nearby, we can exploit the fact that the groups addressed by a single command are usually neighbors. We do this by defining \emph{barriers} for multicast, such that $barrier(G_{send},G_{recv})$ = $t$ means that the group $G_{send}$ promised that no commands with timestamp lower than $t$ would be sent to group $G_{recv}$ anymore. When a server $S$ has received all the barrier values from its neighbours, and they are all greater than a value $t$, then $S$ knows that no more commands are coming from other groups and that, once the local ordering is done, all the commands with timestamp up to $t$ can be conservatively delivered. Besides, a barrier is sent along with the bundle of all messages with timestamp greater than the last previous barrier sent from $G_{send}$ to $G_{recv}$, so that when a server has received a barrier from a group, it means that it knows all the messages sent by that group until the time value stored in that barrier.

@@

  In this case, when $C$ is optimistically delivered by the coordinator, it starts a Paxos consensus instance within its group, having not only its own group's servers as learners, but also those of the neighbor groups affected by $C$. The proposed value is the command $C$ itself, including $C.pTS$. When the value is decided and learnt, the servers of every other affected group $G'$ adjust $barrier(G, G')$ to $C.pTS$ and the coordinators of all the involved groups insert it into the respective \textit{consPending} lists.

@@

A more formal description of the protocol is given in Algorithm \ref{algorithm:optpaxos}. We consider that three primitives are given: $getTime()$, which returns the current value of the local wallclock; $Propose(val, learners)$, which initiates a Paxos \cite{paxosref} instance proposing the value $val$ with the local group as the group of acceptors and $learners$ as the group of learners; and also $Learn(val)$, which is called when a paxos instance achieves consensus. The $Learn(val)$ is called for all the processes which are supposed to learn a given proposed value once it is decided by a Paxos instance. For the sake of simplicity, we assume that, for each group, the values are learnt by consensus in the same order in which the consensus instances were initiated. This can be easily done with a sequence number given by the group's coordinator when initiating each consensus instance -- but note that we don't care if consensus instances initiated by different groups are finished in a non-deterministic order, but only those initiated in each single group; we want something like a FIFO ``learn order'', and not necessarily atomic.

Moreover, each server contains an \textit{optPending} list which contains the commands waiting to be \textit{OPT-Delivered} (optimistically delivered). The \mbox{\textit{consPending}} list contains the commands ready to be \textit{CONS-Delivered} (conservatively delivered), but which may be waiting for the clearance from the neighbor groups. Also, each command $cmd$ has at least three fields: $dst$, which is the list of groups which contain objects affected by it; $src$, which is the group to which the client who issued it is connected; and $pTS$, the value of the wallclock of the server when it received the command from the client. Note that, by abuse of notation, we have `server $s \in C.dst$' instead of `server $s$~: $\exists$ group $G~\in~C.dst$ $\wedge$ $s~\in~G$.

\begin{algorithm}
\begin{distribalgo}[1]
\STATE \underline{This part of the algorithm is executed by every server $s$:}
\STATE
\INDENT{\textit{When $s$ receives a command $cmd$ from a client}}
  \STATE $cmd.pTS \leftarrow getTime()$  
  \COMMENT{current wallclock value as the timestamp of $cmd$}
  \INDENT{\textbf{for all} $s' \in cmd.dst$ \textbf{do}}
    \STATE $send(s', cmd)$    
    \COMMENT{send optimistically $cmd$ to all involved servers}
  \ENDINDENT
  \STATE $optPending \leftarrow optPending$ $\cup$ $\{cmd\}$
\ENDINDENT
\STATE
\INDENT{\textit{When a server $s$ receives a command $cmd'$ from another server}}
  \IF {$cmd'.pTS < getTime() - w(s)$}
    \STATE discard $cmd'$
    \COMMENT{late commands probably lead to out-of-order delivery}
  \ELSE
    \STATE $optPending \leftarrow optPending$ $\cup$ $\{cmd'\}$
  \ENDIF
\ENDINDENT
\STATE
\INDENT{\textit{When $\exists c \in optPending$ : $getTime() > c.pTS + w(s)$ $\wedge$\\$\nexists c' \in optPending: c'.pTS < c.pTS$}}
  \STATE OPT-Deliver(c)  
  \INDENT{\textbf{for all} $cli \in clients(s)$ \textbf{do}}
    \STATE $send(cli, \{c, \text{`optimistic'}\})$    
    \COMMENT{s sends $c$ to its clients, warning it is optimistic}
  \ENDINDENT
  \STATE $optPending \leftarrow optPending \setminus \{c\}$
  \IF {$s$ is a coordinator $\wedge$ $|c.dst| = 1$}
    \STATE $consPending \leftarrow consPending$ $\cup$ $\{c\}$
  \ENDIF  
\ENDINDENT
\STATE
\INDENT{\textit{When} $Learn(\{c, \text{`deliver'}\})$}
  \STATE CONS-Deliver(c)
  \INDENT{\textbf{for all} $cli \in clients(s)$ \textbf{do}}
    \STATE $send(cli, \{c, \text{`final'}\})$    
    \COMMENT{s sends $c$ to its clients, telling it is the final order}
  \ENDINDENT
  \IF {$s$ is a coordinator}
    \STATE $consPending \leftarrow consPending \setminus \{c\}$
  \ENDIF  
\ENDINDENT
\STATE
\STATE
\STATE \underline{The next part is executed only by the coordinator $s_c$ of each group $g$:}
\STATE
\INDENT{\textit{When $\exists c \in optPending$ : $getTime() > c.pTS + w(s_c)$ $\wedge$ \\ $|c.dst| > 1 \wedge c.src = g \wedge \nexists c' \in optPending : c'.pTS < c.pTS$}}
  \STATE $Propose(\{c, \text{`multicast'}\}, c.dst)$
  \COMMENT{saving c in the acceptors before multicasting}
\ENDINDENT
\STATE
\INDENT{\textit{When} $Learn(\{c, \text{`multicast'}\})$}
  \STATE $consPending \leftarrow consPending$ $\cup$ $\{c\}$ 
  \IF {$c.src \neq g$}
    \STATE $barrier(c.src, g) \leftarrow c.pTS$
  \ENDIF  
\ENDINDENT
\STATE
\INDENT{\textit{When $\exists c \in consPending$ : \\ $c.pTS < getTime() - w(s_c)$ $\wedge$ \\ $\forall g' : g'$ is a neighbor of $g$, $c.pTS < barrier(g',g)$ $\wedge$ \\ $\nexists c' \in consPending : c'.pTS < c.pTS$ }}
  \STATE $Propose(\{c, \text{`deliver'}\}, g)$
\ENDINDENT

@@

The problem with Algorithm \ref{algorithm:optpaxos} is that it does not guarantee liveness when a group has no message to receive from some other group and then keeps waiting for a new message to increase the barrier value and proceed with the conservative delivery. However, it does not disrupt the optimistic delivery and liveness for the conservative delivery can be easily provided by sending periodic empty commands to neighbors to which no message has been sent for a specified time threshold \textit{barrierThreshold}. Algorithm \ref{algorithm:liveness} describes this.

\begin{algorithm}
\begin{distribalgo}[1]
\STATE \underline{This algorithm is executed only by the coordinator $s_c$ of each group $g$:}
\STATE
\INDENT{\textit{When} $Learn(\{c, \text{`multicast'}\})$ $\wedge$ $c.src = g$}
  \INDENT{\textbf{for all} $g' : g' \in c.dst$ \textbf{do}}
    \STATE $lastBarrierSent(g') \leftarrow c.pTS$
  \ENDINDENT
\ENDINDENT
\STATE
\INDENT{\textit{When $\exists g' : g'$ is a neighbor of $g$ $\wedge$\\$getTime() - lastBarrierSent(g') > barrierThreshold$}}
  \STATE $ec \leftarrow$ empty command
  \STATE $ec.pTS \leftarrow getTime()$
  \STATE $ec.src \leftarrow g$
  \STATE $Propose(\{ec, \text{`multicast'}\}, \{g,g'\})$
  \COMMENT{saving that nothing was sent until ec.pTS}
\ENDINDENT
\STATE

\caption{Achieving liveness}
\label{algorithm:liveness}
\end{distribalgo}
\end{algorithm}
